<!-- for comments -->
  <style>
  div.comment pre { background-color:#ffbbbb; }
  div.comment pre.r { background-color:#ffbbbb; }
  div.comment {color: red;}
  </style>
---
title: "WCR 11"
author: "Natalie Martin"
date: "2024-11-13"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgbif)
library(tidyverse)
library(MuMIn)
library(data.table)
library(ggmap)
library(usmap)
library(magick)
library(cowplot)
library(lme4)
library(car)
library(rnoaa)
library(dplyr)
```

#1 
```{r}
snow <- read_delim("https://bcorgbio.github.io/class/data/snow.txt")
```

#2
```{r}
yr <- c(2001,2002,2003,2004,2005,2006,2007,2008,2009,2010,2011,2012,2013,2014,2015,2016,2017,2018,2019,2020,2021)

#CPK: or yr <- 2001:2021
yw_l <- list()

for (i in yr){
  yw_l[[i]] <- occ_data(scientificName = "Acanthis flammea",
               year=i,
               limit=10000,
               country="US",
               basisOfRecord = "HUMAN_OBSERVATION",
               stateProvince = "Maine")[[2]] %>% 
    select(individualCount,year,month,day,decimalLongitude,decimalLatitude)
}

cr <- do.call(rbind,yw_l)
```

#3
```{r}
yr_count <- cr %>% 
  group_by(year) %>% 
  summarise(count=sum(individualCount, na.rm=T)) #CPK: Yes! na.rm
```

#4
```{r}
cr_snow <- left_join(yr_count,snow,by="year")
```
#5
```{r}
cr_snow %>% 
  ggplot(aes(x=snow_in,y=count,col=year))+geom_point()
```
#Interpretation
In more recent years we have seen more Acanthis flammea. Moderate snow totals, between 50 and 100 inches, typically correspond to higher Acanthis flammea observation rates. 



<div class="comment">
Excellent work! 

Total points: 10
</div>
